---
title: "Data Analysis with R and the tidyverse"
subtitle: "Lecture 1"
author: "John Thompson"
institute: "University of Leicester"
date: "21st August 2022"
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css, my_theme.css]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: true
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(xaringanthemer)
style_duo_accent(
  primary_color = "#4d90c4",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
xaringanExtra::use_tile_view()
```

## Vectors

collections of numbers, characters or booleans
.panelset[
.panel[.panel-name[creation]
```{r}
v <- c(2.3, 4.6, 1.9)
print(v)
```

```{r}
v <- c("Jack", "Ali", "Margaret", "Jane")
print(v)
```
]
.panel[.panel-name[In one step]
```{r eval=FALSE}
y <- fun3(fun2(fun1(x)))
```

Difficult to read and error prone
]
.panel[.panel-name[Pipe]
```{r eval=FALSE}
x %>%
  fun1() %>%
  fun2() %>%
  fun3()   -> y
```

Each returned object is fed (piped) into the next function

Most R users code the pipe as
```{r eval=FALSE}
y <- x %>%
       fun1() %>%
       fun2() %>%
       fun3()
```
]
.panel[.panel-name[First argument]
The piped object becomes the **first argument** of the next function  

```{r eval=FALSE}
x %>%
  fun1(y, z)  ->  w
```

is the same as  
```{r eval=FALSE}
w <- fun1(x, y, z)
```
]
.panel[.panel-name[The dot]
You can force a different position using **the dot**

```{r eval=FALSE}
x %>%
  fun1(y, ., z)  ->  w
```

is the same as  

```{r eval=FALSE}
w <- fun1(y, x, z)
```

]
]
---
## Data frames and tibbles

A **data frame** is a rectangular data object  

- rows represent observations  
- cols represent variables  

A **tibble** is the tidyverse version of a data frame
```{r}
tibble( id     = 1:4,
        age    = c(39, 45, 22, 56),
        gender = c("m", "f", "f", "m") ) %>%
  print()
```

---
## Reading an Excel file


.panelset[
.panel[.panel-name[read_excel()]
Use the **read_excel()** function from the **readxl** package

```{r eval=FALSE}
# load the package
library(readxl)

# read the Excel file
myDF <- read_excel(path = "C:/Projects/icu/data/myFile.xlsx") 
```
]
.panel[.panel-name[help]
Obtain help on any available function using *?functionName*

```{r eval=FALSE}
?read_excel
```

```{r eval=FALSE}
read_excel(
  path,
  sheet = NULL,
  range = NULL,
  col_names = TRUE,
  col_types = NULL,
  na = "",
  trim_ws = TRUE,
  skip = 0,
  n_max = Inf,
  guess_max = min(1000, n_max),
  progress = readxl_progress(),
  .name_repair = "unique"
)
```
]
.panel[.panel-name[Quick and Dirty]

When argument names are not specified, the arguments are assigned in order.

It is common to see,
```{r eval=FALSE}
myDF <- read_excel("C:/Projects/icu/data/myFile.xlsx") 
```

Best practice is to specify the argument names. (principle of literate programming)
]
]
---
## Data Cleaning

The **janitor** package contains functions for cleaning a tibble, including  

- **clean_names()** creates column names in a consistent style  
- **remove_empty()** removes blank rows and/or columns  
- **remove_constant()** removes columns when every value is the same  

---
## Data Cleaning example

.panelset[
.panel[.panel-name[Messy Excel]
<img src = "C:/Projects/RCourse/ISCB/site/figs/messy_data.jpg" />

]
.panel[.panel-name[read_excel()]
```{r warning=FALSE, message=FALSE}
library(readxl)

read_excel("../data/messy_data.xlsx") %>%
  print()
```

missing values denoted by **NA**  
two word names enclosed in back ticks, e.g. **`patient age`**
]
.panel[.panel-name[read and clean]
```{r warning=FALSE, message=FALSE}
library(janitor)

read_excel("../data/messy_data.xlsx") %>%
  clean_names(case = "lower_camel") %>%
  remove_empty() %>%
  print()
```

**case="snake"** is a popular option, **patient_age**
]
]
---
## Saving in rds format

.panelset[
.panel[.panel-name[rds]
- Save the cleaned data  
- Will not need to clean it again  
- re-cleaning would invite inconsistencies  

**rds** is a fast, compact way of saving data  

Save with **write_rds()** and read back with **read_rds()**
]
.panel[.panel-name[Example]
```{r eval=FALSE}
# read the messy data
read_excel("data/rawData/messy_data.xlsx") %>%
  # clean the data
  clean_names(case = "lower_camel") %>%
  remove_empty() %>%
  # save in rds format
  write_rds("data/rData/clean_data.rds")
```

```{r eval=FALSE}
# read and print the clean data
cleanDF <- read_rds("data/rData/clean_data.rds")

print(cleanDF)
```
]
]
---
## Data manipulation

**dplyr** is a tidyverse package for manipulating data in tibbles.

The functions are designed to work with the pipe.

The most important functions are,  

- **rename()**: change the name of a column (variable)    
- **select()**: choose a subset of the columns (variables)  
- **filter()**: choose a subset of the rows (subjects)  
- **mutate()**: alter a column (variable) by calculation or create a new column   
- **group_by()**: group together the rows (subjects) into subsets   
- **summarise()**: calculate summary statistics   

---
## Filter  

subset of the rows

.panelset[
.panel[.panel-name[Raw Data]
```{r}
cleanDF <- read_rds("../data/clean_data.rds")

print(cleanDF)
```
]
.panel[.panel-name[filter sex]
```{r}
cleanDF %>%
  filter( sex == "F" ) %>%
  print()
```
]
.panel[.panel-name[filter age]
```{r}
cleanDF %>%
  filter( patientAge > 40 ) %>%
  print()
```
]
]
---
## Select or Rename Variables

.panelset[
.panel[.panel-name[Raw Data]
```{r}
cleanDF <- read_rds("../data/clean_data.rds")

print(cleanDF)
```
]
.panel[.panel-name[select]
```{r}
cleanDF %>%
  filter( sex == "M" ) %>%
  select( systolic, diastolic ) %>%
  print()
```
]
.panel[.panel-name[rename]
```{r}
cleanDF %>%
  rename( age    = patientAge,
          gender = sex,
          sbp    = systolic,
          dbp    = diastolic) %>%
  print()
```
]
]
---
## Summarise and Group by

.panelset[
.panel[.panel-name[Raw Data]
```{r}
cleanDF <- read_rds("../data/clean_data.rds")

print(cleanDF)
```
]
.panel[.panel-name[summarise]
```{r}
cleanDF %>%
  summarise( n         = n(),
             sbpMean   = mean(systolic),
             sbpStd    = sd(systolic) ) %>%
  print()
```
]
.panel[.panel-name[group_by]
```{r}
cleanDF %>%
  group_by( sex ) %>%
  summarise( n         = n(),
             sbpMean   = mean(systolic),
             sbpStd    = sd(systolic) ) %>%
  print()
```
]
]

---
## Mutate

.panelset[
.panel[.panel-name[Raw Data]
```{r}
cleanDF <- read_rds("../data/clean_data.rds")

print(cleanDF)
```
]
.panel[.panel-name[mutate]
```{r}
# calculate the mean arterial pressure
cleanDF %>%
  mutate( map = (systolic + 2 * diastolic) / 3  ) %>%
  print()
```
]
.panel[.panel-name[combination]
```{r}
cleanDF %>%
  rename( sbp = systolic) %>%
  select( id, sbp) %>%
  mutate( mSBP      = mean(sbp),
          deviation = sbp - mSBP) %>%
  print()
```
]
]
---
## Visualisation

**ggplot2** is the package of functions for graphics  

Key principles of ggplot2

- plots are build up in layers  
- the layers are joined by a **+**  
- when options are not specified explicitly, their values are inherited from the previous layer   

Terminology  

- aspects of the plot that **depend on the data** are specified with the aesthetic function, `aes()`  
- the contents of the plot (lines, points, bars etc.) are controlled by functions with names that start with `geom_`  
---
## Scatter plots
.panelset[
.panel[.panel-name[Raw Data]
```{r}
cleanDF <- read_rds("../data/clean_data.rds")

print(cleanDF)
```
]
.panel[.panel-name[simple]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic)) +
  geom_point()
```
]
.panel[.panel-name[colour & size]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic, colour = sex)) +
  geom_point( size = 3)
```
]
.panel[.panel-name[shape & labels]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic, colour = sex)) +
  geom_point( size = 3, shape = 17 ) +
  labs(x     = "Age (years)", 
       y     = "DBP", 
       title = "Diastolic Blood Pressure by Patient's age")
```
]
.panel[.panel-name[swap]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic, shape = sex)) +
  geom_point( size = 3, colour = "blue" ) +
  labs(x     = "Age (years)", 
       y     = "DBP", 
       title = "Diastolic Blood Pressure by Patient's age")
```
]
]
---
## Smooths

.panelset[
.panel[.panel-name[lowess]
```{r out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic)) +
  geom_point() + 
  geom_smooth()
```
]
.panel[.panel-name[regression]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic)) +
  geom_point() + 
  geom_smooth( method="lm")
```
]
]
---
## Facets

.panelset[
.panel[.panel-name[columns]
```{r out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic)) +
  geom_point() + 
  geom_smooth( method="lm") +
  facet_grid( . ~ sex) 
```
]
.panel[.panel-name[rows]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  ggplot( aes(x = patientAge, y = diastolic)) +
  geom_point() + 
  geom_smooth( method="lm") +
  facet_grid( sex ~ .) 
```
]
]
---
## Histogram

.panelset[
.panel[.panel-name[histogram]
```{r out.width = '600px',  fig.align='center'}
cleanDF %>%
  mutate( map = (systolic + 2 * diastolic) / 3  ) %>%
  ggplot( aes(x = map ) ) +
    geom_histogram( bins = 5, fill = "steelblue" ) +
    labs(x     = "MAP", 
         title = "Mean Arterial Pressure")
```
]
.panel[.panel-name[density]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  mutate( map = (systolic + 2 * diastolic) / 3  ) %>%
  ggplot( aes(x = map ) ) +
    geom_density( bins = 5, fill = "steelblue" ) +
    labs(x     = "MAP", 
         title = "Mean Arterial Pressure")
```
]
]
---
## Tests in a pipe

.panelset[
.panel[.panel-name[t-test]
```{r out.width = '600px',  fig.align='center'}
cleanDF %>%
  t.test(systolic ~ sex, data = .) %>%
  print()
```
]
.panel[.panel-name[Wilcoxon rank sum]
```{r  out.width = '600px',  fig.align='center'}
cleanDF %>%
  wilcox.test(systolic ~ sex, data = .) %>%
  print()
```
]
]
---
## rmarkdown

**markdown**, a simple markup language, plus **chunks** of R code.  

**YAML** header controls the document style   

RStudio contains many short-cuts to aid report production.  
