---
title: "Data Analysis with R and the tidyverse"
subtitle: "Introduction"
author: "John Thompson"
institute: "University of Leicester"
date: "21st August 2022"
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css, my_theme.css]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: true
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#4d90c4",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
xaringanExtra::use_tile_view()
```

class: left, middle

.pull-left[

### Statistician & Genetic Epidemiologist
<br>
### john.thompson@le.ac.uk
<br>
### Blog on Teaching R
### https://staffblogs.le.ac.uk/teachingr/
]

.pull-right[
<img src=figs/john02.jpg width=350, height=600>
]
---
## The Design of the R language

- Designed in 1970s by John Chambers specifically for statistical data analysis  

- Implemented first as the commercial product S and then as the free version R  

- First stable release in 2000, now with over 10 million users  

- Key design feature is R's **flexibility**  

- Cannot teach R without considering how it will be used  

- This is an opinionated course; it links R to a style of data analysis  

---
## Functional programming

All computer languages work with  

- **objects**: structured data, e.g. numbers, matrices, strings etc.  

- **functions**: code that manipulates the data objects  

Languages that emphasise objects are called object orientated (OO) and those that emphasise the functions are all functional programming (FP) languages.  

R is flexible so can be used in both style. I will emphasise the FP approach.  

- This course only covers the use of R existing functions.  
- The next stage is to write your own functions.  
- Perhaps, 90% of all data analysis uses existing functions.  
- Your own functions are needed for developing new methods of analysis and to make you code more efficient.  
---
## A typical piece of R Code

```{r eval=FALSE}
# create a data object
evenTimes <- c(12, 5, 4, 1, 7, 5, 3)

# use a function to create a new data object
meanTime <- mean(eventTimes)

# use a function to print the new object
print(meanTime, digits=3)
```

Objects are assigned to their names using `<-`  
function names are always followed by brackets, e.g. c(), mean()
The inputs to the function, called arguments, are placed in the brackets and separated by commas
`=` is used for assigning values to arguments  
<br>
Set of related functions are collected together in **packages**.  

Often the most difficult task is to find a function that does what you want. There are 100's of base R functions and 100,000s of functions downloadable from the internet.
---
## The tidyverse  

Hadley Wickham sees data analysis in stages  

<center>
<img src=figs/workflow.png>
</center>

He, with others, have written packages of functions for each stage in an analysis  
- **readr** and **haven** for importing data  
- **dplyr** and **tidyr** for manipulating data  
- **ggplot2** for plotting data  
- **broom** for summarising models  
- **rmarkdown** for report writing  

There are about 20 packages, known collectively as the **tidyverse**
---
## A Style of Data Analysis

My style is guided by three main principles  

- **Reproducibility**: you must be able to reproduce all of your results exactly  
- **Literate programming**: the scripts that you write should be easy to follow  
- **Structured workflows**: you should work in the same way on all of your data analysis projects  

Following these principles will **increase efficiency** and **decrease errors**.  

I will teach my personal style. You don't need to follow my style, but you do need a style.

For example, here are two styles aimed at producing literate code  

- https://google.github.io/styleguide/Rguide.html  
- http://style.tidyverse.org/  

My coding style is different, but I try to be consistent.

---
## Checklist: Prior to starting the analysis 
 
- Select a short, descriptive project name   

- Create a project folder with the same name as the project    

- Create your standard folder structure within the project folder. Here is the structure that I use,  

&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;data  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;dataStore  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;rawData    
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;rData  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;docs  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;R  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;rmd  
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;temp  


- Create a text file called **diary.txt** in the docs subfolder    

- Copy the original data files to the rawData folder   

- Open RStudio and create an RStudio project that points to your project folder. Start the diary.
---
## Checklist: During the Data Analysis  

- Never work interactively. Interactive analysis is not reproducible.    

- Document everything as you go along: its tiresome but essential. Put comments in the scripts and notes in the diary.    

- Never edit the original data files - Instead, read the raw data files into R, process them and then save the processed data for later use.  

- Write separate R scripts for each stage in the analysis (**targets**)  

- Never duplicate information in your data files - the risk of inconsistencies is too great   

- Keep a backup copy of your work    

- Archive the data and all of your scripts whenever you reach an important milestone, such as the publication of a report (**renv** and **git**)  






