---
title: "Data Analysis with R and the tidyverse"
subtitle: "Lecture 2"
author: "John Thompson"
institute: "University of Leicester"
date: "21st August 2022"
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css, my_theme.css]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: true
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(xaringanthemer)
style_duo_accent(
  primary_color = "#4d90c4",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
xaringanExtra::use_tile_view()
```

## Vectors

collections of numbers, characters or booleans
.panelset[
.panel[.panel-name[creation]
```{r}
v <- c(2.3, 4.6, 1.9)
print(v)
```

```{r}
v <- c("Jack", "Ali", "Margaret", "Jane")
print(v)
```
]
.panel[.panel-name[Square brackets]
```{r}
v[3]
```

```{r}
v[c(1,4)]
```
]
]
---
## Tidy data

<br>

- The `tidyverse` works best when data are stored in `tidy format`. 

- https://www.jstatsoft.org/article/view/v059i10. 

- similar to the format used in relational databases.  

- arrange data into tables (data frames or tibbles in R) so that,

  - each row represents a different subject or observation    
  - each column represents a different variable or measurement    
  - data are not duplicated in different tables

---
## Wide and Long formats

.panelset[
.panel[.panel-name[wide]
```{r}
tibble( id           = 1:4,
        age          = c(37, 23, 62, 55),
        sbp_baseline = c(135, 120, 155, 140),
        sbp_followup = c(125, 115, 140, 135)) %>%
  print() -> wideDF
```
]
.panel[.panel-name[Long]
```{r}
tibble( id    = rep(1:4, each=2),
        visit = rep( c("baseline", "followup"), times=4),
        age   = rep( c(37, 23, 62, 55), each=2),
        sbp   = c(135, 125, 120, 115, 155, 140, 140, 135)) %>%
  print() -> longDF
```
]
.panel[.panel-name[Long to Wide]
```{r}
longDF %>%
   pivot_wider( names_from  = visit, 
                values_from = sbp   ) %>%
   print()
```
]
.panel[.panel-name[Wide to Long]
```{r}
wideDF %>%
  pivot_longer( cols      = c(sbp_baseline, sbp_followup), 
                names_to  = "visit", 
                values_to = "sbp") %>%
  print()
```
]
.panel[.panel-name[neater]
```{r}
wideDF %>%
  pivot_longer( cols      = c(sbp_baseline, sbp_followup), 
                names_to  = "visit", 
                values_to = "sbp") %>%
  mutate( visit = str_replace(visit, "sbp_", "")) %>%
  print()
```
]
]
---
## Missing values

.panelset[
.panel[.panel-name[data]
```{r}
tibble( id  = 1:5,
        sex = c("m", "m", "f", "f", "f"),
        age = c( 45,  32,  57,  63,  23),
        sbp = c(125, 130,  NA, 150, 135)) %>%
  print()   -> sbpDF
```
]
.panel[.panel-name[locate missing]
```{r}
sbpDF %>%
  mutate( missing  = is.na(sbp),
          measured = !is.na(sbp)) %>%
  print()
```
]
.panel[.panel-name[filter not missing]
```{r}
sbpDF %>%
  filter( !is.na(sbp)) %>%
  print()
```
]
]
---
## Calculation with NA

.panelset[
.panel[.panel-name[data]
```{r}
tibble( id  = 1:5,
        sex = c("m", "m", "f", "f", "f"),
        age = c( 45,  32,  57,  63,  23),
        sbp = c(125, 130,  NA, 150, 135)) %>%
  print()   -> sbpDF
```
]
.panel[.panel-name[stats]
```{r}
sbpDF %>%
  summarise( n    = n(),
             mAge = mean(age),
             mSBP = mean(sbp)) %>%
  print()
```
]
.panel[.panel-name[filter]
```{r}
sbpDF %>%
  filter( !is.na(sbp)) %>%
  summarise( n    = n(),
             mAge = mean(age),
             mSBP = mean(sbp)) %>%
  print()
```
]
.panel[.panel-name[na.rm]
```{r}
sbpDF %>%
  summarise( n    = n(),
             mAge = mean(age, na.rm = TRUE),
             mSBP = mean(sbp, na.rm = TRUE)) %>%
  print()
```
]]
---
## Factors

Variables with a finite set of possible values

.panelset[
.panel[.panel-name[data]
```{r}
sbpDF %>%
  str()
```
]
.panel[.panel-name[to factor]
```{r}
sbpDF %>%
  mutate( sex = factor(sex)) %>%
  str() 
```
]
.panel[.panel-name[terminology]
Internally all factors are stored as numbers, 1, 2, etc.

- **levels** ... values/words taken by the variable in the dataset  
- **labels** ... values/words that you want to appear in R output  

default: levels assigned alphabetically, labels=levels
]
.panel[.panel-name[full specification]
```{r}
# --- Convert sex from character to factor ----------------------
sbpDF %>%
  mutate( sex = factor(sex, 
                       levels = c("m", "f"),
                       labels = c("male", "female"))) %>%
  str()
```
]
]
---
## flextable

package **flextable** contains many functions for producing report quality tables

it is not the only option.  

`gt`, `huxtable`, `xtable` and `ztable` and many other packages do a similar job.

table creation has two stages  

- create a tibble containing the table contents  

- use **flextable** to add headers and control the layout  

---
## flextable example 

.panelset[
.panel[.panel-name[data]
```{r}
tibble( id  = 1:5,
        sex = c("m", "m", "f", "f", "f"),
        age = c( 45,  32,  57,  63,  23),
        sbp = c(125, 130,  NA, 150, 135)) %>%
  print()   -> sbpDF
```
]
.panel[.panel-name[contents]
```{r}
sbpDF %>%
  mutate( sex = factor(sex, 
                       levels = c("m", "f"),
                       labels = c("male", "female"))) %>%
  group_by(sex) %>%
  summarise( meanAge = mean(age, na.rm=TRUE),
             meanSBP = mean(sbp, na.rm=TRUE)) %>%
  print()  -> tabDF
```
]
.panel[.panel-name[default]
```{r}
library(flextable)

tabDF %>%
  flextable()
```
]
.panel[.panel-name[prettier]
```{r eval=FALSE}
tabDF %>%
  flextable() %>%
  border_outer() %>%
  set_header_labels(
    meanAge = "mean age",
    meanSBP = "mean SBP"
  ) %>%
  set_caption("Average age and SBP of mean and women") %>%
  colformat_double(
    j      = c( 2, 3),
    digits = 1
  )
```
]
.panel[.panel-name[output]
```{r echo=FALSE}
tabDF %>%
  flextable() %>%
  border_outer() %>%
  set_header_labels(
    meanAge = "mean age",
    meanSBP = "mean SBP"
  ) %>%
  set_caption("Average age and SBP of mean and women") %>%
  colformat_double(
    j      = c( 2, 3),
    digits = 1
  )
```
]
.panel[.panel-name[refs]
<br>
- on-line book https://ardata-fr.github.io/flextable-book/  
<br>
- list of functions https://davidgohel.github.io/flextable/reference/index.html  
<br>
- gallery https://ardata.fr/en/flextable-gallery/  
]
]
---
## ggplot themes

set of defaults for ggplot2  

.panelset[
.panel[.panel-name[standard]
```{r out.width = '600px',  fig.align='center', warning = FALSE}
sbpDF %>%
  ggplot( aes(x = age, y = sbp)) +
  geom_point( size = 3 )
```
]
.panel[.panel-name[theme_light]
```{r out.width = '600px',  fig.align='center', warning = FALSE}
sbpDF %>%
  ggplot( aes(x = age, y = sbp)) +
  geom_point( size = 3 ) + 
  theme_light()
```
]
.panel[.panel-name[theme_dark]
```{r out.width = '600px',  fig.align='center', warning = FALSE}
sbpDF %>%
  ggplot( aes(x = age, y = sbp)) +
  geom_point( size = 3 ) + 
  theme_dark()
```
]
.panel[.panel-name[ggthemes]
```{r out.width = '600px',  fig.align='center', warning = FALSE}
library(ggthemes)

sbpDF %>%
  ggplot( aes(x = age, y = sbp)) +
  geom_point( size = 3 ) +
  theme_stata()
```
]
]
---
## Adapting the theme

.panelset[
.panel[.panel-name[theme_classic]
```{r out.width = '600px',  fig.align='center', warning = FALSE}
sbpDF %>%
  mutate( sex = factor(sex, 
                       levels = c("m", "f"),
                       labels = c("male", "female"))) %>%
  ggplot( aes(x = age, y = sbp, colour = sex)) +
  geom_point( size = 3 ) +
  theme_classic()
```
]
.panel[.panel-name[adapt_code]
```{r out.width = '600px',  fig.align='center', warning = FALSE, eval=FALSE}
sbpDF %>%
  mutate( sex = factor(sex, 
                       levels = c("m", "f"),
                       labels = c("male", "female"))) %>%
  ggplot( aes(x = age, y = sbp, colour = sex)) +
  geom_point( size = 3 ) +
  scale_colour_manual(values = c("male" = "darkgreen", "female" = "orange")) +
  theme_classic() +
  theme( legend.position = c(0.1, 0.8) )
```
]
.panel[.panel-name[adapt_output]
```{r out.width = '600px', out.height = '400px',  fig.align='center', warning = FALSE, echo=FALSE}
sbpDF %>%
  mutate( sex = factor(sex, 
                       levels = c("m", "f"),
                       labels = c("male", "female"))) %>%
  ggplot( aes(x = age, y = sbp, colour = sex)) +
  geom_point( size = 3 ) +
  scale_colour_manual(values = c("male" = "darkgreen", "female" = "orange")) +
  theme_classic() +
  theme( legend.position = c(0.1, 0.8) )
```
]
]

---
## Error Bars

.panelset[
.panel[.panel-name[Long]
```{r}
tibble( id    = rep(1:4, each=2),
        visit = rep( c("baseline", "followup"), times=4),
        age   = rep( c(37, 23, 62, 55), each=2),
        sbp   = c(135, 125, 120, 115, 155, 140, 140, 135)) %>%
  print() -> longDF
```
]
.panel[.panel-name[stats]
```{r}
longDF %>%
  group_by(visit) %>%
  summarise( n = n(),
             m = mean(sbp),
             s = sd(sbp))  %>%
  mutate( se = s / sqrt(n),
          lb = m - 1.96 * se,
          ub = m + 1.96 * se) %>%
  print() -> statDF
```
]
.panel[.panel-name[error bars code]
```{r eval=FALSE}
statDF %>%
  ggplot( aes(x=visit, y=m)) +
  geom_point( size=4) +
  geom_errorbar( aes(ymin=lb, ymax=ub), width=0.2) +
  labs( y = "mean SBP") +
  theme_light()
```
]
.panel[.panel-name[plot]
```{r out.width = '600px', out.height = '400px',  fig.align='center', warning = FALSE, echo=FALSE}
statDF %>%
  ggplot( aes(x=visit, y=m)) +
  geom_point( size=4) +
  geom_errorbar( aes(ymin=lb, ymax=ub), width=0.2) +
  labs( y = "mean SBP") +
  theme_light()
```
]
]
