---
title: "What Next"
subtitle: "Data Analysis with R and the tidyverse"
author: "John Thompson"
institute: "University of Leicester"
date: "21st August 2022"
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css, my_theme.css]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: true
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(xaringanthemer)
style_duo_accent(
  primary_color = "#4d90c4",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
xaringanExtra::use_tile_view()
```

## Milestones in learning the tidyverse

### Introductory  

Use of the tidyverse packages and the pipe without writing your own functions  

### Intermediate

Adding user-written functions for efficiency, code organisation and coding new methods. Working with list columns in a tibble.  

shiny for dashboards and interactive graphics  

### Advanced

Larger projects: analysis workflows, version control, collaboration, databases, package development, parallel processing, the two language problem

---
## Introductory: Sessions 2 and 3

### Tidy Format

Organising data in rectangular tables  

long and wide tables, conversion with **pivot_wider()** and **pivot_longer()**

joining tables, with **left_join()**, **right_join()**, **full_join()**, etc.  

### Factors

defining levels and labels, 

### Pretty tables

the **flextable** package

---
## Introductory: Sessions 2 and 3

### More ggplot2 

themes, colours, error bars, plotting factors


### Missing values

their impact on computations

### Statistical models

base R model fitting with a pipe  

the **broom** package  

---
## Introductory: Sessions 2 and 3

### Working with lists  

list as collections of data objects  

extracting items from a list by name (**$**) or by position (**[[ ]]**)   

saving metadata as attributes

### Programming tools

repeating a calculation in a loop, accumulating results  

conditional computation with **ifelse()**  

blocks of code in a pipe with **{ }**,  more on the dot in a pipe
---
## Intermediate: writing a function

writing your own function  

```{r}
# ---------------------------------------------
# function for the width of an 80% interval
# x is a numeric vector (not checked)
#
range80 <- function(x) {
  lb <- quantile(x, prob = 0.1)
  ub <- quantile(x, prob = 0.9)
  range <- ub - lb
  return(range)
}
```

---
## Intermediate: using your own function

using your own function  

```{r}
tibble( id    = 1:8,
        age   = c(37, 23, 62, 55, 42, 71, 19, 38),
        sbp   = c(135, 125, 120, 115, 155, 140, 140, 135)) %>%
  summarise( n = n(),
             m = mean(sbp),
             r = range80(sbp) ) %>%
  print()
```

---
## Repeating a calculation

finding the 80% range of more than one column  

```{r}
tibble( id    = 1:8,
        age   = c(37, 23, 62, 55, 42, 71, 19, 38),
        sbp   = c(135, 125, 120, 115, 155, 140, 140, 135)) %>%
  summarise( across( age:sbp, range80)) %>%
  print()
```

---
## Replacing loops

This code comes from session 3 of the introductory course calculates 100 bootstrap medians

```{r eval=FALSE}
tibble( id    = 1:8,
        age   = c(37, 23, 62, 55, 42, 71, 19, 38),
        sbp   = c(135, 125, 120, 115, 155, 140, 140, 135)) -> sbpDF

# data frame for saving 100 bootstrap medians
tibble( index  = 1:100,
        median = rep(NA, 100)) %>%
  print() -> resultsDF

# seed for reproducibility
set.seed(982)

# repeat the calculation 100 times
for( i in 1:100 ) {
  sbpDF %>%
    slice_sample(n = 8, replace = TRUE) %>%
    summarise( m = median(sbp)) %>%
    { .$m }  -> resultsDF$median[i]
}
```

---
## Repeating without loops

place the contents of the loop in a function and use **rerun()**

.panelset[
.panel[.panel-name[setup]
```{r out.width = '600px',  fig.align='center', warning = FALSE}
# the data
tibble( id    = 1:8,
        age   = c(37, 23, 62, 55, 42, 71, 19, 38),
        sbp   = c(135, 125, 120, 115, 155, 140, 140, 135)) -> sbpDF

# the function
boot_median <- function(df) {
  df %>%
    slice_sample(n = 8, replace = TRUE) %>%
    summarise( m = median(sbp)) %>%
    { .$m }  %>%
    return()
}
```
]
.panel[.panel-name[repeat in a pipe]
```{r out.width = '600px',  fig.align='center', warning = FALSE}
# seed for reproducibility
set.seed(982)

# tibble containing 3 bootstrap medians
rerun(5, boot_median(sbpDF) ) %>%
  flatten_dbl() %>%
  as_tibble_col("median") %>%
  mutate( index = 1:5)
```
]
]

---
## the purrr package

**purrr** has many functions for repeating computations  

**map()**, **map2()**, **pmap()**, etc.  

<br>

- no need for loops  
- no need to break the flow of a piped analysis when repeating a computation  
- testing is easier  
- code is shorter and easier to read  

---
## list columns

.panelset[
.panel[.panel-name[data]
```{r}
tibble( id  = 1:8,
        sex = c("m", "m", "f", "f", "f", "m", "f", "m"),
        age   = c(37, 23, 62, 55, 42, 71, 19, 38),
        sbp   = c(135, 125, 120, 115, 155, 140, 140, 135)) %>%
  mutate( sex = factor(sex, levels = c("m", "f"),
                       labels = c("male", "female"))) %>%
  print()   -> sbpDF
```
]
.panel[.panel-name[nest]
```{r}
sbpDF %>%
  group_by( sex ) %>%
  nest() 
```
]
.panel[.panel-name[fit models]
```{r}
# function to fit a model
myModel <- function(df) {
  model = lm( sbp ~ age, data = df)
  return(model)
}

sbpDF %>%
  group_by( sex ) %>%
  nest() %>%
  mutate( model = map(data, myModel))
```
]
.panel[.panel-name[unnest]
```{r}
library(broom)

sbpDF %>%
  group_by( sex ) %>%
  nest() %>%
  mutate( model = map(data, myModel)) %>%
  mutate( coef  = map(model, tidy)) %>%
  select( sex, coef) %>%
  unnest( coef)
```
]
]
---
## Shiny dashboards

### A gallery of examples https://shiny.rstudio.com/gallery/

---
## Advanced Topics

Methods associated to scaling up to larger projects, e.g. funded research, PhD, collaborative research, commercial data analysis

### Analysis Workflows

large projects have many components that feed into one another. The **targets** package keeps track of the different components and how they interact.  

---
## Advanced Topics

### Version control

**git** keep tracks of the changes to your code as it develops and enables you to revert to earlier versions.

The **renv** package enable you to control the version number of the packages that you use.  

### Collaboration

**github** enables the sharing of code and integrates with **git**. Any of the many cloud data repositories will enable data sharing.  

---
## Advanced Topics

### Databases

Efficient ways of storing large amounts of data. **duckdb** stores data by variable and is good for data analysis. It integrates well into R.  

### Parallel Processing

Important for computationally intensive analyses, such as MCMC. The **future** package. **furrr** is a parallel version of **purrr**.  

---
## Advanced Topics
 ### Package development

Collecting your functions into packages. Initially for internal use, but perhaps also for wider distribution. Documentation and testing.

### Two language problem

R is easy to use and flexible, but it is relatively slow. Behind the scenes, the tidyverse makes use of a lot of C code. **Rcpp** is a package that enables C++ code to be integrated into R.  


