---
title: "Data Analysis with R and the tidyverse"
subtitle: "Demonstration 1"
author: "John Thompson"
institute: "University of Leicester"
date: "21st August 2022"
output:
  xaringan::moon_reader:
    css: [xaringan-themer.css, my_theme.css]
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: "16:9"
      countIncrementalSlides: true
---

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(xaringanthemer)
style_duo_accent(
  primary_color = "#4d90c4",
  secondary_color = "#FF961C",
  inverse_header_color = "#FFFFFF"
)
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
  fig.width=9, fig.height=3.5, fig.retina=3,
  out.width = "100%",
  cache = FALSE,
  echo = TRUE,
  message = FALSE, 
  warning = FALSE,
  hiline = TRUE
)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
xaringanExtra::use_tile_view()
```

## Demonstration

Data from

Iesu, E., Franchi, F., Zama Cavicchi, F., Pozzebon, S., Fontana, V., Mendoza, M., ... & Taccone, F. S. (2018).  
**Acute liver dysfunction after cardiac arrest.**  
PLoS One, 13(11), e0206655..   

- Department of Intensive Care at Erasme Hospital in Brussels. 
- all patients who had a cardiac arrest either in or out of hospital between 2007 and 2015
- exclude people who died within 24 hours of their arrest 
- Outcome for such patients is poor 

**analysis of the number of days spent in ICU**

---
## Extra functions

- **glimpse()**  - summarise the contents of a data frame or tibble  
- **is.na()**    - return TRUE if missing (NA) and FALSE otherwise   
- **!**          - negation or NOT, turns TRUE to FALSE & FALSE to TRUE  
- **!is.na()**   - return TRUE if NOT missing  
- **median()**   - **sum()**, **min()**, **max()** - return median, total, minimum, maximum   
- **quantile()** - return a quantile, i.e. value a proportion prob of the way through the sorted values, median(x) is the same as quantile(x, prob=0.5) **kruskal.test()**  
- Kruskal-Wallis nonparametric test for difference in location between 2 or more groups.  
- **scale_y_log10()** - in ggplot, use a log10 scale for the y-axis. Also scale_x_log10(), scale_x_log(), scale_x_reverse() and many others   
- **subtitle** - an extra argument for ggplot labs()  

---
## Script header

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#==================================================================
# File:        demo1.R
# Project:     icu (project folder C:/Projects/ISCB/icu)
# Author:      john thompson
# Date:        22 July 2022
# Description: Analysis of length of Stay in ICU 
#==================================================================
library(tidyverse)
library(readxl)
library(janitor)

excel_file <- "C:/Projects/RCourse/ISCB/data/rawData/iesu/Table 1 Enrica.xlsx"
clean_file <- "C:/Projects/Rcourse/ISCB/data/rData/cardiac.rds"
```
]
.panel[.panel-name[output]

no output for this code
```{r echo=FALSE, message=FALSE, warning=FALSE}
#==================================================================
# File:        demo1.R
# Project:     icu (project folder C:/Projects/ISCB/icu)
# Author:      john thompson
# Date:        22 July 2022
# Description: Analysis of length of Stay in ICU 
#==================================================================
library(tidyverse)
library(readxl)
library(janitor)

excel_file <- "C:/Projects/RCourse/ISCB/data/rawData/iesu/Table 1 Enrica.xlsx"
clean_file <- "C:/Projects/Rcourse/ISCB/data/rData/cardiac.rds"
```
]
]
---
## Read and clean

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Read & clean the data
#
read_excel( excel_file ) %>%
  remove_empty() %>%
  remove_constant() %>%
  clean_names( "lower_camel" ) %>%
  filter( !is.na(id) ) %>%
  write_rds( clean_file )

#-----------------------------------------------------------------
# Read clean data
#
cardiacDF <- read_rds( clean_file )
```
]
.panel[.panel-name[output]

no output for this code
```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# Read & clean the data
#
read_excel( excel_file ) %>%
  remove_empty() %>%
  remove_constant() %>%
  clean_names( "lower_camel" ) %>%
  filter( !is.na(id) ) %>%
  write_rds( clean_file )

#-----------------------------------------------------------------
# Read clean data
#
cardiacDF <- read_rds( clean_file )
```
]
]
---
## Inspect the data

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Data summary
#
cardiacDF %>%
  glimpse()
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# Data summary
#
cardiacDF %>%
  glimpse()
```
]
]
---
## mean and std of ICU stay

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Summarise ICU stay
#
cardiacDF %>%
  summarise( n       = n(),
             icuMean = mean(icuStay),
             icuStd  = sd(icuStay)) 
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# Summarise ICU stay
#
cardiacDF %>%
  summarise( n       = n(),
             icuMean = mean(icuStay),
             icuStd  = sd(icuStay)) 
```
]
]
---
## Histogram of ICU stay

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Histogram of Days in ICU 
#
cardiacDF %>%
  ggplot( aes(x = icuStay)) +
  geom_histogram( bins = 50, fill = "steelblue" ) +
  labs( x     = "Days",
        title = "Number of days spent in ICU following a Cardiac Arrest") 
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width = '600px', out.height = '450px', fig.align='center'}
#-----------------------------------------------------------------
# Histogram of Days in ICU 
#
cardiacDF %>%
  ggplot( aes(x = icuStay)) +
  geom_histogram( bins = 50, fill = "steelblue" ) +
  labs( x     = "Days",
        title = "Number of days spent in ICU following a Cardiac Arrest") 
```
]
]
---
## Quantiles of ICU stay

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Summarise ICU stay
#
cardiacDF %>%
  summarise( n         = n(),
             icuMin    = min(icuStay),
             icuQ1     = quantile(icuStay, prob=0.25),
             icuQ3     = quantile(icuStay, prob=0.75),
             icuMax    = max(icuStay) )
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# Summarise ICU stay
#
cardiacDF %>%
  summarise( n         = n(),
             icuMin    = min(icuStay),
             icuQ1     = quantile(icuStay, prob=0.25),
             icuQ3     = quantile(icuStay, prob=0.75),
             icuMax    = max(icuStay) )
```
]
]
---
## Quantiles by Mortality

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Mortality & ICU stay
#
cardiacDF %>%
  group_by( hospDeath ) %>%
  summarise( n         = n(),
             icuMedian = median(icuStay),
             icuQ1     = quantile(icuStay, prob=0.25),
             icuQ3     = quantile(icuStay, prob=0.75)) %>%
  mutate( pct = 100 * n / sum(n) ) %>%
  select( hospDeath, n, pct, icuMedian, icuQ1, icuQ3)
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# Mortality & ICU stay
#
cardiacDF %>%
  group_by( hospDeath ) %>%
  summarise( n         = n(),
             icuMedian = median(icuStay),
             icuQ1     = quantile(icuStay, prob=0.25),
             icuQ3     = quantile(icuStay, prob=0.75)) %>%
  mutate( pct = 100 * n / sum(n) ) %>%
  select( hospDeath, n, pct, icuMedian, icuQ1, icuQ3)
```
]
]
---
## Histograms by Mortality

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Days in ICU by Death in Hospital
#
cardiacDF %>%
  ggplot( aes(x = icuStay)) +
  geom_histogram( bins = 50, fill = "steelblue" ) +
  labs( x        = "Days in ICU",
        title    = "Length of ICU Stay by Death in Hospital",
        subtitle = "Shorter stay in those that die") +
  facet_grid( hospDeath ~ . )
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width = '600px', out.height = '450px', fig.align='center'}
#-----------------------------------------------------------------
# Days in ICU by Death in Hospital
#
cardiacDF %>%
  ggplot( aes(x = icuStay)) +
  geom_histogram( bins = 50, fill = "steelblue" ) +
  labs( x        = "Days in ICU",
        title    = "Length of ICU Stay by Death in Hospital",
        subtitle = "Shorter stay in those that die") +
  facet_grid( hospDeath ~ . )
```
]
]
---
## Test of ICU stay by Mortality

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Wilcoxon test of Days in ICU by Death in Hospital
#
cardiacDF %>%
  wilcox.test( icuStay ~ hospDeath, data = .)
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# Wilcoxon test of Days in ICU by Death in Hospital
#
cardiacDF %>%
  wilcox.test( icuStay ~ hospDeath, data = .)
```
]
]
---
## CVD Score and ICU stay

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# CVD score & ICU stay
#
cardiacDF %>%
  mutate( cvd = chronHeartFailure + hypertension + coronArteryDis ) %>%
  group_by( hospDeath, cvd) %>%
  summarise( n         = n(),
             icuMedian = median(icuStay),
             icuQ1     = quantile(icuStay, prob=0.25),
             icuQ3     = quantile(icuStay, prob=0.75))
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# CVD score & ICU stay
#
cardiacDF %>%
  mutate( cvd = chronHeartFailure + hypertension + coronArteryDis ) %>%
  group_by( hospDeath, cvd) %>%
  summarise( n         = n(),
             icuMedian = median(icuStay),
             icuQ1     = quantile(icuStay, prob=0.25),
             icuQ3     = quantile(icuStay, prob=0.75))
```
]
]
---
## Analysis of Survivors: ICU stay and cvd

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# cvd and days in ICU for survivors
#
cardiacDF %>%
  filter( hospDeath == 0 ) %>%
  mutate( cvd = chronHeartFailure + hypertension + coronArteryDis ) %>%
  ggplot( aes(x = icuStay)) +
  geom_histogram( bins = 50, fill = "steelblue" ) +
  labs( x        = "Days in ICU",
        title    = "Length of ICU Stay by CVD score in patients that survive",
        subtitle = "no clear association with CVD score") +
  facet_grid( cvd ~ . ) 
```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width = '600px', out.height = '450px', fig.align='center'}
#-----------------------------------------------------------------
# cvd and days in ICU for survivors
#
cardiacDF %>%
  filter( hospDeath == 0 ) %>%
  mutate( cvd = chronHeartFailure + hypertension + coronArteryDis ) %>%
  ggplot( aes(x = icuStay)) +
  geom_histogram( bins = 50, fill = "steelblue" ) +
  labs( x        = "Days in ICU",
        title    = "Length of ICU Stay by CVD score in patients that survive",
        subtitle = "no clear association with CVD score") +
  facet_grid( cvd ~ . ) 
```
]
]
---
## Analysis of Survivors: test of cvd

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# Kruskal-Wallis test of Days in ICU for survivors
#
cardiacDF %>%
  filter( hospDeath == 0 ) %>%
  mutate( cvd = chronHeartFailure + hypertension + coronArteryDis ) %>%
  kruskal.test( icuStay ~ cvd, data = .)

```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE}
#-----------------------------------------------------------------
# Kruskal-Wallis test of Days in ICU for survivors
#
cardiacDF %>%
  filter( hospDeath == 0 ) %>%
  mutate( cvd = chronHeartFailure + hypertension + coronArteryDis ) %>%
  kruskal.test( icuStay ~ cvd, data = .)
```
]
]
---
## ICU stay and RoSC

.panelset[
.panel[.panel-name[code]

```{r eval=FALSE}
#-----------------------------------------------------------------
# ICU Stay & Time to Return of Spontaneous Circulation (RoSC)
#
cardiacDF %>%
  ggplot( aes(x = timeToRoscMin, y = icuStay) ) +
  geom_point( colour = "darkgreen") +
  scale_y_log10() +
  geom_smooth( method = "lm" , colour = "black") +
  labs( x        = "Time to Spontaneous Circulation (mins)",
        y        = "Days in ICU on log scale",
        title    = "Length of ICU stay by time to ROSC faceted by death in hospital",
        subtitle = "evidence of an interaction") +
  facet_grid( hospDeath ~ .) 

```
]
.panel[.panel-name[output]

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width = '600px', out.height = '450px', fig.align='center'}
#-----------------------------------------------------------------
# ICU Stay & Time to Return of Spontaneous Circulation (RoSC)
#
cardiacDF %>%
  ggplot( aes(x = timeToRoscMin, y = icuStay) ) +
  geom_point( colour = "darkgreen") +
  scale_y_log10() +
  geom_smooth( method = "lm" , colour = "black") +
  labs( x        = "Time to Spontaneous Circulation (mins)",
        y        = "Days in ICU on log scale",
        title    = "Length of ICU stay by time to ROSC faceted by death in hospital",
        subtitle = "evidence of an interaction") +
  facet_grid( hospDeath ~ .) 

```
]
]

